\documentclass[11pt,a4paper]{article}

\usepackage[margin=2.45cm,headheight=23.12pt]{geometry}
\usepackage[T1]{fontenc}
\usepackage{tgheros}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{setspace}
\usepackage{parskip}
\usepackage{fancyhdr}
\usepackage{booktabs}
\usepackage{array}
\usepackage{lastpage}
\usepackage[hidelinks]{hyperref}

\renewcommand{\familydefault}{\sfdefault}
\definecolor{headinggrey}{gray}{0.22}
\definecolor{rulegrey}{gray}{0.72}
\definecolor{tableheadgrey}{gray}{0.92}

\setstretch{1.08}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.8em}
\emergencystretch=2em

\pagestyle{fancy}
\fancyhf{}
\lhead{ENEL445 -- Applied Engineering Optimisation}
\rhead{Transcript Corrections}
\cfoot{\thepage\ of \pageref{LastPage}}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\headrule}{\hbox to\headwidth{\color{rulegrey}\leaders\hrule height \headrulewidth\hfill}}

\begin{document}

\vspace*{0.5em}

{\raggedleft
David Ewing\\
Student ID: 82171165\\
ENEL445 -- Applied Engineering Optimisation\\
University of Canterbury\\[0.3em]
5 May 2026\par
}

\vspace{1.2em}

Professor Roy S.\ Smith\\
Department of Information Technology and Electrical Engineering\\
ETH Z\"urich\\[0.3em]
\textit{(Erskine Visiting Fellow, University of Canterbury)}

\vspace{1.8em}

\textbf{Subject: Corrections to the Least Squares Regression Lecture Transcript}

\vspace{0.4em}
\noindent\textcolor{rulegrey}{\rule{\textwidth}{0.4pt}}
\vspace{0.8em}

Dear Professor Smith,

I hope this message finds you well. I have been reviewing the transcript of your lecture on least squares regression and have identified a number of errors --- some arising from voice-recognition transcription artefacts, and others appearing to be factual inaccuracies when checked against standard historical sources. I have compiled them below for your information.

I should note that the corrections to historical details were verified against the Wikipedia articles for Carl Friedrich Gauss, Ceres (dwarf planet), and Franz Xaver von Zach. I recognise that these are secondary sources and that you may have access to more authoritative references; I offer the corrections in good faith and welcome any further clarification.

\vspace{0.6em}
\textbf{\large Voice-Recognition Transcription Errors}
\vspace{0.3em}

The following appear to be artefacts of automatic speech recognition, where words or names were misheard:

\renewcommand{\arraystretch}{1.35}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.42\textwidth}
                >{\raggedright\arraybackslash}p{0.42\textwidth}}
\toprule
\textbf{As transcribed} & \textbf{Corrected to} \\
\midrule
\textit{right essential, right essential angle} & right ascension angle \\
\textit{xeris} & Ceres \\
\textit{over what happened to 44 days} & over what amounted to 44 days \\
\textit{1805 genre published a paper sort of on the method of the method of} & 1805 Legendre published a paper sort of on the method of \\
\textit{conic sectors} (three instances) & conic sections \\
\textit{con acceptors} & conic sections \\
\textit{conic sections, conic sectors} (stray doublet) & conic sections \\
\textit{matrix calculus, calculus.org} & \texttt{matrixcalculus.org} \\
\textit{from et cetera} & from ETH \\
\bottomrule
\end{tabular}

\vspace{0.6em}
\textbf{\large Attribution Error}
\vspace{0.3em}

The transcript read: \textit{``within half a degree of where Zach said it would be in that but Zach, but Gauss, would not say.''}

This appeared to be a garbled version of the intended meaning. I corrected it to: \textit{``within half a degree of where Gauss said it would be, but Gauss would not say,''} since it was Gauss who made the prediction, not Zach (Zach was the astronomer who published the observational data and subsequently recovered Ceres).

\vspace{0.6em}
\textbf{\large Expansion of First Mention of Zach}
\vspace{0.3em}

The first reference to ``Zach'' gave no indication of who he was. I added a brief identification:
\textit{``Zach --- Franz Xaver von Zach, editor of the \textit{Monatliche Correspondenz zur Bef\"orderung der Erd- und Himmels-Kunde} (Monthly Correspondence for the Promotion of Earth and Sky Science). He published the observational data,''}
so that readers unfamiliar with the history have sufficient context.

\vspace{0.6em}
\textbf{\large Historical Factual Corrections}
\vspace{0.3em}

The following corrections were made on the basis of historical sources:

\renewcommand{\arraystretch}{1.35}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.42\textwidth}
                >{\raggedright\arraybackslash}p{0.42\textwidth}}
\toprule
\textbf{As transcribed} & \textbf{Corrected to} \\
\midrule
\textit{Titus Bode law} & Titius--Bode law \\
 & (after Johann Daniel Titius, not ``Titus'') \\[0.3em]
\textit{22 observations} / \textit{And he made 22 of them} & 24 observations / And he made 24 of them \\
 & (Piazzi made 24 observations of Ceres before losing it) \\[0.3em]
\textit{seventh of December} & 31st of December \\
 & (Ceres was recovered by Zach on 31 December 1801; he also observed on 7 December but the confirmed recovery is attributed to the 31st) \\[0.3em]
\textit{five years earlier} & ten years earlier \\
 & (Gauss claimed to have been using the method since 1794 or 1795, approximately ten years before Legendre published in 1805) \\
\bottomrule
\end{tabular}

\vspace{1.2em}

I hope these corrections are useful. Please do not hesitate to let me know if any of my changes deviate from your original intent, or if you have a preferred phrasing for any of the passages.

Thank you for an engaging and illuminating lecture.

\vspace{1.8em}

Yours sincerely,

\vspace{1.4em}

David Ewing\\
ENEL445, University of Canterbury

\end{document}
